---
title: "Example Dashboard"
author: You! 
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: united
---

```{r setup, include=FALSE}
library(flexdashboard)
```

# Basic Analysis

Column {data-width=650}
-----------------------------------------------------------------------

```{r}
library(tidyverse)
library(knitr)
df <- read_csv('../data/listings.csv')

top_rated_neighborhood <- df %>% 
  group_by(neighbourhood) %>% 
  summarise(review_scores_rating = mean(review_scores_rating, na.rm = T)) %>% 
  arrange(desc(review_scores_rating)) %>% 
  filter(neighbourhood != 'NA')
```
### Histogram of Ratings

```{r}
df %>% 
  ggplot() + 
  aes(x = review_scores_rating) + 
  geom_histogram() + 
  theme_bw()
```


Column {data-width=350}
-----------------------------------------------------------------------

### Analysis

There are ```r nrow(top_rated_neighborhood)``` neighborhoods represented in the data set. Of these, the highest rated is ```r top_rated_neighborhood$neighbourhood[1]``` and the lowest is ```r top_rated_neighborhood$neighbourhood[nrow(top_rated_neighborhood)]```. This data is current as of ```r lubridate::today()```

### Top Rated Neighbourhoods
```{r}
top_rated_neighborhood %>% 
  head(5) %>% 
  kable()
```

# Interactivity Example

Column {data-width=650}
-----------------------------------------------------------------------

### Map of Listing Locations

```{r}
library(leaflet)
m <- leaflet(df) %>% addTiles()
m %>% 
  addCircleMarkers(radius = 1)
```

Column {data-width=350}
-----------------------------------------------------------------------

### Leaflet

Leaflet is an interactive visualization platform based on `D3.js`. It is possible to construct Leaflet visualizations directly, but the `leaflet` package provides a convenient way to  build Leaflet visualizations from your data. Leaflet and many other interesting HTML-based widgets are 100% compatible with RMarkdown in all formats, making it easy to embed interesting interactive graphics into your work. 
